\chapter*{Conclusion}

In this thesis we introduced algorithms for solving strong positional games,
mainly PN-search and its enhancements.

We focused on games with high symmetry. The high symmetry allows, by joining
several isomorph game state into one, to narrow the partial game tree. When
partial game tree is narrow, we can search in it deeper. The other feature which
keeps the tree narrow is that we do not deal with high branching factor. For
example, we can compare our game with the game Tzaar, which has high branching
factor and low symmetry. Despite of clever heuristics and optimizations the
best bot playing the game searches the game tree only into depth
4~\cite{paulie}. For comparison, one variant of our solver for (7,4)-clique
game solved the game tree to depth 16.

There are other features distinguishing games with high symmetry. It is much
more important to join the nodes. We must treat carefully the lifetime of
the nodes, that is, we do not want to delete nodes from cache especially when their
solution was expensive because as we may find another ancestor of these nodes.
Finally in games with high symmetry much more often happens situation described
in Figure~\ref{deset}.

\subsubsection{Further work}

There are many enhancements which we did not implemented and even do
not mention. For example 1+$\epsilon$ trick.

It could also be useful to implement enhancement ``dveloping a~node without creating sons''
in Section~\ref{lastenh} and study if it can significantly reduce the partial game
tree. Also to might it may be interesting to try it on other games.

Of course, there is space for more new enhancements and heuristics. 

\addcontentsline{toc}{chapter}{Conclusion}

